TO FIND

### 2013-09-24 submission1/ms.tex
Minor_revisions

%========================================================%
\section*{Abstract} 
The metacommunity perspective appears as an appropriate conceptual framework to
make ecology more predictive. It is particularly relevant to limnology, where
exchanges of organisms and nutrients affect community and ecosystem properties
from the local to the regional scales. The recent development of neutral theory
appears as a step back in that direction because of the assumption of ecological
equivalence and the absence of any effect of the environment on community
organization. A remarkable strength of neutral theory is nonetheless to provide
a general theory of diversity that accounts for a wide range of empirical
observations. In this paper, we argue that neutral theory can be useful to
understand the impact of dispersal on community assembly in landscapes of
various complexities. Our analysis focus on spatially explicit landscapes
conceptualized as networks of local communities (e.g. lakes) connected to each
other by dispersal channels (e.g. rivers). The main objective of the paper is to
use neutral theory to stress the importance of landscape structure on the
distribution of diversity. We refer to the landscape organization as a "spatial
contingency" that could potentially affect the coexistence mechanisms at play.
We briefly review the main approaches to describe spatial networks and describe
three simple toy models of metacommunity dynamics. We take this opportunity to
review their assumptions and main predictions. We then conduct simulations of
these models to reveal with simple examples the impact of spatial network
structure on diversity distribution. The simulation results show that
competitive interactions buffer the potential impact of landscape structure. The
strongest relationship between node position in the landscape and species
richness was observed for the patch dynamics model without any
interactions. On the other hand, strong and unequal competitive interactions
minimized the effect of node position. We conclude that the neutral model is a
useful tool to understand the joint effects of dispersal and ecological
interactions. Our analysis shows that limnologists must now integrate more
realistic landscapes when analyzing community assembly from a metacommunity
perspective.
\newpage

@@

Ecology needs to move toward a more predictive approach, integrating elements of
theoretical ecology \parencite{Thuiller2013}. The metacommunity perspective
\parencite{Leibold2004a} appears naturally as the appropriate conceptual
framework to fill this challenge. The metacommunity concept builds on feedbacks
between local scale processes, such as competitive interactions and local
adaptation, and regional scale processes such as dispersal, gene flow and
speciation. It is particularly relevant to limnology, where exchanges of
organisms and nutrients affect community and ecosystem properties from the local
(e.g. vertical mixing \parencite{Ryabov2011}) to the regional (e.g. connections
of lakes \parencite{Leibold2004b, Muneepeerakul2008, Gravel2010a} scales. The
metacommunity perspective emphasizes the importance of dispersal relative to
pairwise interactions in the organization of ecological communities.

At first sight, the development of neutral theory appears as a step back.
The neutral theory of biodiversity makes the provocative assumption that species are
ecologically equivalent \parencite{Bell2000,Hubbell2001}. Neutral community
dynamics are driven only by demographic stochasticity and dispersal and thus,
variation in the environment has no impact on demography. Neutral theory sparked
a historical debate still lasting after more than a decade
\parencite{Chave2004, Etienne2011, Rosindell2012,Clark2012}. It was stimulated
by the surprising ability of neutral models to fit some well studied empirical
observations such as species abundance distributions and distance-decay
relationships.

A remarkable strength of the theory is to provide a \emph{"formal general theory
of abundance and diversity that will account, in a simple and economical
fashion, for the many patterns that ecologists have documented"}
\parencite{Bell2001}. Even if new studies rejecting neutral theory are
consistently published (e.g. \textcite{Ricklefs2012}, but see the comments
associated with this study, e.g. \textcite{Etienne2012b}), a consensus is
forming that neutral theory is a well-developed null hypothesis for niche theory
and could even be used as an adequate approximation of ecological dynamics in
some situations. \textcite{Bell2001} nicely envisioned two perspectives to
neutral theory that are still standing today. Under the weak perspective,
neutral theory provides a set of realistic predictions of community organization
despite false assumptions. Even if being fundamentally wrong, neutral theory
would still useful when used as a null hypothesis \parencite{Gotelli2006}. It is
considered as an improvement over traditional null hypotheses based on
randomization \parencite{Gotelli2000} because it readily integrates dispersal.
On the other had, the strong version posits that neutral theory is a satisfying
approximation to community dynamics and an appropriate theory to explain the
distribution of biodiversity. It implies that the right mechanisms have been
identified and that the consistently observed differences among species do not
strongly impact community organization.

Neutral theory has also been proposed as a useful tool to understand and predict
some aspects of community dynamics. It links to an old philosophical debate
between realism and instrumentalism \parencite{Wennekes2012}. Because every
ecological model is a simplification of reality, scientists have to subjectively
decide the level of detail they put in, leaving out some elements they consider
unimportant. The realist perspective requires all assumptions of the theory
to be true, while the utility of the theory is more important to
instrumentalism. The utilitarian value of a theory could either be for
understanding or for prediction (another old philosophical debate, see
\textcite{Shmueli2010}). Obviously neutral theory could only be instrumental.
The question then is if such a 'general, large-scale, but vague' theory
\parencite{Wennekes2012} is a satisfying approximation.

The instrumentalist view of neutral theory raises the question of why it should
be a satisfying approximation despite knowing the pieces are wrong? Perhaps
stochasticity of various origins blur the deterministic differences among
species and promote ecological drift \parencite{Gravel2011a}. Much has been said
about the existence of demographic stochasticity \parencite{Clark2012}, and we
will therefore leave this discussion for other papers. A second explanation is that
dispersal and historical contingencies might have a much more profound impact on
species distribution \parencite{Bahn2007, Boulangeat2012} and ecological
dynamics. The debate over the equivalence assumption and demographic
stochasticity might have overlooked the recognition of how much dispersal
influence community assembly.

@@

Our main objective is to use neutral theory to stress the importance of
landscape network structure on the distribution of diversity. We refer to the
landscape organization as a \emph{spatial contingency}
\parencite{Peres-Neto2012} that could potentially affect the coexistence
mechanisms at play. We will therefore move from a perspective where dispersal is
either global or constant over space (e.g. a lattice), to a perspective focusing
on the variance of dispersal. A second generation of neutral models (e.g.
\textcite{Economo2008, Economo2011,Desjardins2012a,Desjardins2012b}), field
surveys \parencite{Muneepeerakul2008} and even experiments
\parencite{Carrara2012}, recently introduced more realistic landscapes and found
surprising contributions of spatial contingencies. We start with a short review
of the main approaches to describe spatial networks. Then we describe three
simple toy models of metacommunity dynamics, using this opportunity to review
their assumptions and main predictions. We provide as Supplementary Material the
R scripts for the toy models and all simulations conducted for this paper. We
then conduct simple simulations of these models to reveal with simple examples
the impact of spatial network structure on diversity distribution. We conclude
with a discussion on the operationality of the metacommunity framework.

@@

The popularity of network theory stems from its ability to model complex
structures while allowing us to extract useful metrics (Table 1). At a very high
level, a network can be described by its number of nodes (the order) and 
edges (the size). Looking more closely, the relationship between
nodes is influenced by paths, which are ordered series of nodes. Centrality
is a \emph{central} concept in network theory, where it can be seen
to as a measure of ``importance''. The simplest measure of the centrality of a
node is its degree, which is the number of nodes directly connected to it. Of
course, this is a very rough description of centrality. For example, two lakes
can have the same degree, with one being connected to a small isolated cluster,
while the other one is part of one of the biggest network of lakes. In this case,
measures of centrality like eigen-centrality will weight the importance of the
connection, so a node connected to well-connected nodes will have higher
centrality than a node connected to isolated nodes.

In this paper we study four types of networks (Fig. 1). First, for the sake of
comparisons, we study a complete network where dispersal is global. Then we
consider a lattice, which is the most regular type of spatially explicit
network. Then after we simulated two types of random spatial networks. We
considered random geometric graphs \parencite{sed01}. In this algorithm, all
nodes are assigned to a position in some two-dimensional space, most often the
unit square. Then, all pairs of nodes within some threshold Euclidean distance
$r$ are connected with an edge. The resulting networks have the desirable
property of locality: if a node $A$ is connected to two vertices $B$ and $C$.
then $B$ and $C$ are more likely to be connected than two random vertices.
Random geometric networks have been extensively studied \parencite{app97a,
app97b, app02a, app02b, pen03} and we provide a R function to generate them. The
position of nodes is typically random, but we could also imagine alterations
where they are either more aggregated or segregated than expected by chance
alone. Finally, we also provided the code for a second structure that we call a
random geometric tree. The algorithm first builds a random geometric graph, then
selects a node from which to start the tree. It then calculates the the shortest
path tree \parencite{dij59} from this node to all other ones and remove edges
not located along this tree. This random geometric tree does not exactly
represent dendritic landscapes but is a convenient model to simulate a lake
connected by rivers to a series of smaller lakes.

@@

In this section we describe three toy models representing different
perspectives of metacommunity ecology: patch dynamics, neutral dynamics and
species sorting. While the neutral model is interesting in itself, it is by its
comparison with a model without any interactions (patch dynamics) and with niche
differentiation (species sorting) that we will be able to fully understand the
interaction between these processes and landscape structure. Despite being neutral,
competitive interactions in neutral models are very strong because of the
zero-sum assumption (the community is always at carrying capacity). We will
first review the fundamental assumptions of each model with their description
(Table 2 summarizes the parameters and variables that are used), and then
briefly discuss their main predictions. Simulation results are presented in the
next section, with the corresponding R code provided in the Supplementary
Material.

@@

The simplest metacommunity model is a $S$ species extension of traditional
metapopulation models \parencite{Hanski1999}. The standard Levins metapopulation
model \parencite{Levins1969} describes the stochastic colonizations and
extinctions of a single species over a homogenous landscape. The basic unit is
the population. The model tracts the dynamics of occupancy (the fraction of the
landscape that is occupied) with an ordinary differential equation and therefore
assumes an infinite landscape. The simulation model we run is more realistic as
it simulates a finite number $N$ of discrete patches (or nodes in network
terminology). The rules described in the previous section were used to generate
connectivity matrices along four scenarios (Fig. 1): global dispersal (connected
graph), a lattice, a random geometric graph and a random tree graph. A patch $x$
shares $d_x$ links with neigbhouring patches (its degree). At each time step
(the simulation model is discrete in time), the probability that a colonist
coming from an occupied patch $y$ arrives at patch $x$ is $cd_y^{-1}$, where $c$
is the probability a colonization event takes place if all connected patches are
occupied. The expected probability that a colonist arrives to patch $x$ from
patch $y$ is then $C_{ixy}=cp_{iy}d_y^{-1}$, where $p_{i}y$ is the probability
that patch $y$ is occupied by species $i$. The probability that an extinction
occurs in a given patch is $e$. The Levins model is for a single species, but a
basic metacommunity patch dynamics model could be run by aggregating $S$
independent metapopulation models \parencite{Hanski1997}. There are no
interactions in this simple model, which means there is no limit to local
species richness and no carrying capacity. Competitive, mutualistic and
predator-prey interactions have been added to this framework (e.g.
\textcite{Tilman1994a,Holt1996,Klausmeier1998,Gravel2011b}) but we will keep this
model minimal for the sake of comparison with the neutral model.

Predictions of the patch dynamics metacommunity model are quite straightforward.
First, a fundamental result of metapopulation ecology is that persistence will
occur if colonization probability is larger than extinction probability ($c>e$).
Given that all species are the same, then we should expect the regional
diversity ($\gamma$) to be $S$ if this condition is satisfied and $0$ if not.
The situation is however more complex in spatially explicit landscapes with
complex connectivity matrices \parencite{Hanski1998}. Spatially explicit
dispersal usually reduces the occupancy and thereby the likelihood of
persistence. The second prediction is that, given spatial variation in
connectivity, there will be spatial variation in occurrence probability. Given
the above formulation of a colonization event to occur, the probability that an
empty location is colonized by an offspring coming from a least one of the $r$ neighbouring patches is $I_{ix}=1-\prod_{y=1}^r d_x(1-C_{ixy})$. This equation
basically tells us that the colonization probability will increase
asymptotically with the degree of a patch (because of the product). It is easy
to show from metapopulation theory that the occurrence probability in a patch is
then $p_{ix}=I_{ix}(I_{ix}+e)^{-1}$. The feedback between local and regional dynamics
arises because all $p_{ix}$ from the landscape are dependent from each other.
Simulations are usually conducted to solve the model for a large landscape, but
numerical solutions are theoretically possible. The aggregation across the $S$
species of the regional species pool is obtained by taking the summation of
occurrence probabilities over all species, $s_x = \sum{p_i}$. Because in this
model all species are equal, we expect the local species richness to be a linear
function of the patch degree (number of edges). Multi-species analysis of
metapopulation models also reveals interesting predictions on other aspects of
community organization at various spatial scales such as the species-area
relationship \parencite{Hanski1997}, and proved to be useful in conservation
ecology with predictions of extinctions following habitat destruction
\parencite{Tilman1994b,Rybicki2013}.

@@

Neutral theory introduces strong competitive interactions by assuming there is a
finite number of individuals that could occupy a patch. There are different ways
to simulate this \emph{zero-sum rule} \parencite{Bell2000,Hubbell2001}, but they
all result in the same constraint that the increase in abundance of a species
could only occur after an equivalent decrease by another species. One important
change in the formulation of most neutral models relative the patch dynamics
model presented above is therefore that it is individual-based, not population
based. We therefore considered in our toy model of neutral dynamics that each
local patch holds $J_x$ individuals. The model tracks the local abundance of all
species $N_{ix}$ in each local patch. At each time step an individidual dies
with probability $k$. Recruitment only occurs in vacant sites, similarly to a
tree by tree replacement process in a closed canopy forest.

The formulation of the recruitment probability is the central piece of all
neutral models, making the coupling with the metacommunity and
neighbouring patches possible. We adopt a simple formulation based on
\parencite{Gravel2006}. The approach is conceptually similar to placing a trap
in a canopy gap and picking a seed at random among the ones falling in to
determine the identity of the recruited species. The composition of the seed
pool in that trap will be a mixture of local dispersal and immigrants from the
metacommunity. For simplicity, we consider three spatial scales of dispersal but
it would be easy to generalize the approach to a continuous seed dispersal
kernel \parencite{Gravel2006}. The parameter $m$ is the probability that the
recruit is a migrant from neighbouring patches, $M$ is the probability it comes
from a larger (and fixed) metacommunity, and consequently, by substraction,
$1-m-M$ is the probability it comes from local dispersal. The fraction
$N_{ix}J_x^{-1}$ is the local relative abundance and $P_{ix}$ is the relative
abundance of species $i$ in the seed pool coming from patches connected to patch $x$.
The relative abundance in the neighborhood is weighted by the degree of the
connected nodes because some nodes will spread their seeds across a higher
number of nodes and thus contribute less to the seed pool. We thus consider
$P_{ix} =\frac{\sum P_{iy}d_y^{-1}}{\sum d_y^{-1}}$. We assume for simplicity
that the relative abundance in the metacommunity is uniform, i.e. equal to
$S^{-1}$, but other distributions could be used. The metacommunity is usually
modelled with a log-series (e.g. \textcite{Hubbell2001}), but it has been shown
that for a range of realistic dispersal rates a uniform distribution does not
have distinguishable impacts relative to a log-series \parencite{Bell2000}. This
immigration prevents the collapse of the metacommunity to a single species,
since otherwise all species except one will face extinction by ecological drift
(speciation prevents this phenomenon to occur in \textcite{Hubbell2001}). The
local recruitment probability $Pr_{ix}$ is consequently $Pr_{ix} = MS^{-1} +
mP_{ix} + (1-m-M)N_{ix}J_x^{-1}$. Self-replacement occurs when an individual of
species $i$ is replaced by an offspring of the same species coming either from
the remaining local individuals or from the metacommunity.

The model is neutral because it assumes that the probabilities of local
recruitment, immigration and mortality events are all equal across species.
Demographic stochasticity is the source of variations in abundance, but larger
disturbances could be simulated as well, as long as they hit all species with
the same probability, independently of their identity. The fundamental feature
of neutral dynamics is therefore the ecological drift, defined as population
changes emerging from neutraly stable population dynamics. It can be measured as
the variance between replicated time series of community dynamics
\parencite{Gravel2011a}. \textcite{Hubbell2001} provides a very comprehensive
analysis of the model, with specific attention to the effect of the different
parameters on drift (and consequently variance in abundance) and time to
extinction. Despite its simplicity, the neutral model is surprisingly rich in
the predictions it makes. \textcite{Bell2001} and \textcite{Hubbell2001}
analyzed the performance of neutral models to predict species abundance
distributions, the range-abundance relationship, spatial variation in abundance,
the species-area relationship, community turnover (beta-diversity) and
co-occurrence. Recent trophic neutral models were also found to predict
realistic ecological network structures \parencite{Canard2012}. Other than the
ecological equivalence assumption, one of the most criticized aspect of neutral
models is the realism of the speciation process and the required speciation
rates to sustain species richness \parencite{Ricklefs2003,Etienne2007}. Recent
neutral models with more credible speciation models
\parencite{Rosindell2010,Desjardins2012a} revealed the difficulty to maintain
diversity in neutral models over macro-evolutionary time scales. These models
nonetheless proposed interesting predictions on endemic species richness and
island biogeography \parencite{Rosindell2011,Desjardins2012b}.

@@

In this section we provide simple simulation results to illustrate the impact of
spatial contingencies on species distribution and coexistence. We consider four
different landscapes, illustrated at Fig. 1. with the outcome of simulations
using the neutral model. All of these networks have the same number of nodes,
but differ both in number of edges (e.g. dispersal routes between sampling
sites) and connectivity between nodes. We ask how these differences
in topological structures will shape the emerging properties of the community under the
scenarios represented by each metacommunity model. Our analysis is not
exhaustive, it is provided simply to illustrate the interaction between
metacommunity perspectives and landscape structures on $\alpha$, $\beta$ and
$\gamma$ diversity.

In Fig. 2, we present the species richness of each node of the network ($\alpha$
diversity), as a function of the centrality of the node, under different
assumptions of metacommunity dynamics and network structure. We scaled the
species richness by the maximal $\alpha$ diversity to facilitate comparison
between models. The model parameterization is responsible for differences in
both $\alpha$ and $\gamma$ diversity, meaning that only the shape of the
relationship between centrality and richness ought to be looked at. It appears
that both in the random geographical and tree networks, the path dynamics model has a
much more considerable variation in $\alpha$ diversity. However, in all
cases the $\alpha$ diversity increases with the node degree centrality, meaning
that nodes with more connections also host a more diverse community.
Eigen-centrality gave a less clear-cut result, which can probably be
attributed to the fact that our networks are relatively small in size.
Eigen-centrality reports how well your neighbors are connected, and in graphs
with a short diameter (i.e. the two farthest points are not extremely far
apart), this measure might hold less information.

>>>>

%========================================================%
\section*{Abstract} 
The metacommunity perspective appears as an appropriate
conceptual framework to make ecology more predictive. The recent development of
neutral theory appears as a step back in that direction because of the
assumption of ecological equivalence and the absence of any effect of the
environment on community organization. A remarkable strength of neutral theory
is nonetheless to provide a general theory of diversity that accounts for a wide
range of empirical observations. In this paper, we argue that neutral theory can
be useful to understand the impact of dispersal on community assembly in
landscapes of various complexities. Our main objective is to use neutral theory
to stress the importance of complex landscape structure on the
distribution of diversity. We refer to the landscape organization as a "spatial
contingency" that could potentially affect the coexistence mechanisms at play.
We briefly review the main approaches to describe spatial networks and describe
three simple toy models of metacommunity dynamics. We take this opportunity to
review their assumptions and main predictions. We then conduct 
simulations of these models to reveal with simple examples the impact of spatial
network structure on diversity distribution. The simulation results show that
competitive interactions buffer the potential impact of landscape structure. The
strongest centrality-species richness relationship was observed for the patch
dynamics, a model without any interactions. On the other hand, strong and
unequal competitive interactions minimized the effect of centrality. We conclude
that the neutral model is a useful tool to understand the joint effects of
dispersal and community interactions. Our analysis shows that ecologists must
now integrate more realistic landscapes when analyzing community assembly from a
metacommunity perspective. 
\newpage

@@

Ecology needs to move toward a more predictive approach, integrating elements of
theoretical ecology \parencite{Thuiller2013}. The metacommunity perspective
\parencite{Leibold2004a} appears naturally as the appropriate conceptual
framework to fill this challenge. The metacommunity concept builds on feedbacks
between local scale processes, such as competitive interactions and local
adaptation, and regional scale processes such as dispersal, gene flow and
speciation. It is particularly relevant to limnology, where exchanges of
organisms and nutrients affect community and ecosystem properties from the local
(e.g. vertical mixing \parencite{Ryabov2011}) to the regional (e.g. connection
of lakes \parencite{Leibold2004b, Gravel2010a} scales. It emphasizes the
importance of dispersal relative to pairwise interactions in the organization of
ecological communities.

At first sight, the development of neutral theory appears as a step back.
Neutral theory of biodiversity makes the provocative assumption that species are
ecologically equivalent \parencite{Bell2000,Hubbell2001}. Neutral community
dynamics are driven only by demographic stochasticity and dispersal and thus,
variation in the environment has no impact on demography. Neutral theory sparked
an historical debate still lasting after more than a decade
\parencite{Chave2004, Etienne2011, Rosindell2012,Clark2012}. It was stimulated
by the surprising ability of neutral models to fit some well studied empirical
observations such as species abundance distributions and distance-decay
relationships.

A remarkable strength of the theory is to provide a \emph{"formal general theory
of abundance and diversity that will account, in a simple and economical
fashion, for the many patterns that ecologists have documented"}
\parencite{Bell2001}. Even if new studies rejecting neutral theory are
consistently published (e.g. \textcite{Ricklefs2012}), a consensus is forming
that neutral theory is a well-developed null hypothesis for niche theory and
could even be used as an adequate approximation of ecological dynamics in some
situations. \textcite{Bell2001} nicely envisioned two perspectives to neutral
theory that are still standing today. Under the weak perspective, neutral theory
provides a set of realistic predictions of community organization despite false
assumptions. Even if being fundamentally wrong, neutral theory would still
useful when used as a null hypothesis \parencite{Gotelli2006}. It is considered
as an improvement over traditional null hypotheses based on randomization
\parencite{Gotelli2000} because it readily integrates dispersal. On the other
had, the strong version posits that neutral theory is a satisfying approximation
to community dynamics and an appropriate theory to explain the distribution of
biodiversity. It implies that the right mechanisms have been identified and that
the consistently observed differences among species do not impact community
organization.

Neutral theory has also been proposed as a useful tool to understand and predict
some aspects of community dynamics. It links to an old philosophical debate
between realism and instrumentalism \parencite{Wennekes2012}. Because every
ecological model is a simplification of reality, scientists have to subjectively
decide the level of details they put in, leaving out some elements they consider
unimportant. The realist perspective requires all assumptions of the theory
to be true, while the utility of the theory is more important to
instrumentalism. The utilitarian value of a theory could either be for
understanding or for prediction (another old philosophical debate, see
\textcite{Shmueli2010}). Obviously neutral theory could only be instrumental.
The question then is if such a 'general, large-scale, but vague' theory
\parencite{Wennekes2012} is a satisfying approximation.

The instrumentalist view of neutral theory raises the question of why it should
be a satisfying approximation despite knowing the pieces are wrong? Perhaps
stochasticity of various origins blur the deterministic differences among
species and promote ecological drift \parencite{Gravel2011a}. Much has been said
about the existence of demographic stochasticity \parencite{Clark2012}, and we
will therefore let this discussion for other papers. A second explanation is that
dispersal and historical contingencies might have a much more profound impact on
species distribution \parencite{Bahn2007, Boulangeat2012} and ecological
dynamics. The debate over the equivalence assumption and demographic
stochasticity might have overlooked the recognition of how much dispersal
influence community assembly.

@@

Our main objective is to use neutral theory to stress the importance of
landscape network structure on the distribution of diversity. We refer to the
landscape organization as a \emph{spatial contingency}
\parencite{Peres-Neto2012} that could potentially affect the coexistence
mechanisms at play. We will therefore move from a perspective where dispersal is
either global or constant over space (e.g. a lattice), to a perspective focusing
on the variance of dispersal. A second generation of neutral models (e.g.
\textcite{Economo2008, Economo2011,Desjardins2012a,Desjardins2012b}), and even
experiments \parencite{Carrara2012}, recently introduced more realistic
landscapes and found surprising contributions of spatial contingencies. We start
with a short review of the main approaches to describe spatial networks. Then we
describe three simple toy models of metacommunity dynamics, using this
opportunity to review their assumptions and main predictions. We provide as
Supplementary Material the R scripts for the toy models and all simulations
conducted for this paper. We then conduct simple simulations of these models to
reveal with simple examples the impact of spatial network structure on diversity
distribution. We conclude with a discussion on the operationally of the
framework.

@@

The popularity of network theory stems from its ability to model complex
structures while allowing us to extract useful metrics (Table 1). At a very high
level, a network can be described by its number of nodes (the order) and 
edges (the size). Looking more closely, the relationship between
nodes is influenced by paths, which are ordered series of nodes. Centrality
is a \emph{central} concept in network theory, where it can be seen
to as a measure of ``importance''. The simplest measure of the centrality of a
node is its degree, which is the number of nodes directly connected to it. Of
course, this is a very rough description of centrality. For example, two lakes
can have the same degree, with one being connected to a small isolated cluster,
while the other one is part of one of the biggest network of lake. In this case,
measures of centrality like eigen-centrality will weight the importance of the
connection, so a node connected to well-connected nodes will have higher
centrality than a node connected to isolated nodes.

For simulations, spatial networks can easily be generated with the random
geometric graph algorithm \parencite{sed01}. In this algorithm, all nodes are
assigned to a position in some two-dimensional space, most often the unit
square. Then, all pairs of nodes within some threshold Euclidean distance $r$
are connected with an edge. The resulting networks have the desirable property
of locality: if a node $A$ is connected to two vertices $B$ and $C$. then $B$
and $C$ are more likely to be connected than two random vertices. Random
geometric networks have been extensively studied \parencite{app97a, app97b,
app02a, app02b, pen03} and we provide a R function to generate them. The
position of nodes is typically random, but we could also imagine alterations
where they are either more aggregated or segregated than expected by chance
alone.

We also provide the code for a second structure that we call a random geometric
tree. The algorithm first builds a random geometric graph, then select a node
from which to start the tree. It then calculates the the shortest path tree
\parencite{dij59} from this node to all other ones and remove edges not located
along this tree. This random geometric tree does not exactly represent dendritic
landscapes but is a convenient model to simulate a lake connected by rivers to a
series of smaller lakes.

@@

In this section we describe three toy models representing different
perspectives of metacommunity ecology: patch dynamics, neutral dynamics and
species sorting. While the neutral model is interesting in itself, it is by its
comparison with a model without any interactions (patch dynamics) and with niche
differentiation (species sorting) that we will be able to fully understand the
interaction between these processes and landscape structure. Despite neutral,
competitive interactions in neutral models are very strong because of the
zero-sum assumption (the community is always at carrying capacity). We will
first review the fundamental assumptions of each model with their description
(Table 2 summarizes the parameters and variables that are used), and then
briefly discuss their main predictions. Simulation results are presented in the
next section, with the corresponding R code provided in the Supplementary
Material.

@@

The simplest metacommunity model is a $S$ species extension of traditional
metapopulation models \parencite(Hanski1999). The standard Levins metapopulation
model \parencite(Levins1969) describes the stochastic colonizations and
extinctions of a single species over a homogenous landscape. The basic unit is
the population. The model tracts the dynamics of occupancy (the fraction of the
landscape that is occupied) with an ordinary differential equation and therefore
assumes an infinite landscape. The simulation model we run is more realistic as
it simulates a finite number $N$ of discrete patches (or nodes in network
terminology). The rules described in the previous section were used to generate
connectivity matrices along four scenarios (Fig. 1): global dispersal (connected
graph), a lattice, a random geometric graph and a random tree graph. A patch $x$
shares $d_x$ links with neigbhouring patches (its degree). At each time step
(the simulation model is discrete in time), the probability that a colonist
coming from an occupied patch $y$ arrives at patch $x$ is $cd_y^{-1}$, where $c$
is the probability a colonization event takes place if all connected patches are
occupied. The expected probability that a colonist arrives to patch $x$ from
patch $y$ is then $C_{ixy}=cp_{iy}d_y^{-1}$, where $p_{i}y$ is the probability
that patch $y$ is occupied by species $i$. The probability that an extinction
occurs in a given patch is $e$. The Levins model is for a single species, but a
basic metacommunity patch dynamics model could be run by aggregating $S$
independent metapopulation models \parencite(Hanski1997). There are no
interactions in this simple model, which means there is no limit to local
species richness and no carrying capacity. Competitive, mutualistic and
predator-prey interactions have been added to this framework (e.g.
\textcite{Tilman1994a,Holt1996,Klausmeier1998,Gravel2011b}) but we will keep this
model minimal for the sake of comparison with the neutral model.

Predictions of the patch dynamics metacommunity model are quite straightforward.
First, a fundamental result of metapopulation ecology is that persistence will
occur if colonization probability is larger than extinction probability ($c>e$).
Given that all species are the same, then we should expect the regional
diversity ($\gamma$) to be $S$ if this condition is satisfied and $0$ if not.
The situation is however more complex in spatially explicit landscapes with
complex connectivity matrices \parencite{Hanski1998}. Spatially explicit
dispersal usually reduces the occupancy and thereby the likelihood of
persistence. The second prediction is that, given spatial variation in
connectivity, there will be spatial variation in occurrence probability. Given
the above formulation of a colonization event to occur, the probability that an
empty location is colonized is $I_{ix}=1-\prod d_x(1-C_{ixy})$. This equation
basically tells us that the colonization probability will increase
asymptotically with the degree of a patch (because of the product). It is easy
to show from metapopulation theory that the occurrence probability in a patch is
then $p_{ix}=I_x(I_{ix}+e)^{-1}$. The feedback between local and regional dynamics
arises because all $p_{ix}$ from the landscape are dependent from each other.
Simulations are usually conducted to solve the model for a large landscape, but
numerical solutions are theoretically possible. The aggregation across the $S$
species of the regional species pool is obtained by taking the summation of
occurrence probabilities over all species, $s_x = \sum{p_i}$. Because in this
model all species are equal, we expect the local species richness to be a linear
function of the patch degree (number of edges). Multi-species analysis of
metapopulation models also reveals interesting predictions on other aspects of
community organization at various spatial scales such as the species-area
relationship \parencite{Hanski1997}, and proved to be useful in conservation
ecology with predictions of extinctions following habitat destruction
\parencite{Tilman1994b,Rybicki2013}.

@@

Neutral theory introduces strong competitive interactions by assuming there is a
finite number of individuals that could occupy a patch. There are different ways
to simulate this \emph{zero-sum rule} \parencite{Bell2000,Hubbell2001}, but they
all result in the same constraint that the increase in abundance of a species
could only occur after an equivalent decrease by another species. One important
change in the formulation of most neutral models relative the patch dynamics
model presented above is therefore that it is individual-based, not population
based. We therefore considered in our toy model of neutral dynamics that each
local patch holds $J_x$ individuals. The model tracts the local abundance of all
species $N_{ix}$ in each local patch. At each time step an individidual dies
with probability $k$. Recruitment only occurs in vacant sites, similarly to a
tree by tree replacement process in a closed canopy forest.

The formulation of the recruitment probability is the central piece of all
neutral models, making possible the coupling with the metacommunity and
neighbouring patches. We adopt a simple formulation based on
\parencite{Gravel2006}. The approach is conceptually similar to placing a trap
in a canopy gap and picking a seed at random among the ones falling in to
determine the identity of the recruited species. The composition of the seed
pool in that trap will be a mixture of local dispersal and immigrants from the
metacommunity. For simplicity, we consider three spatial scales of dispersal but
it would be easy to generalize the approach to a continuous seed dispersal
kernel \parencite{Gravel2006}. The parameter $m$ is the probability that the
recruit is a migrant from neighbouring patches, $M$ is the probability it comes
from a larger (and fixed) metacommunity, and consequently, by substraction,
$1-m-M$ is the probability it comes from local dispersal. The fraction
$N_{ix}J_x^{-1}$ is the local relative abundance and $P_{ix}$ is the relative
abundance of species $i$ in the seed pool coming from neigbouring patches $x$.
The relative abundance in the neighborhood is weighted by the degree of the
connected nodes because some nodes will spread their seeds across a higher
number of nodes and thus contribute less to the seed pool. We thus consider
$P_{ix} =\frac{\sum P_{iy}d_y^{-1}}{\sum d_y^{-1}}$. We assume for simplicity
(and without loss of generality, \parencite{Bell2000}) that the relative
abundance in the metacommunity is uniform, i.-e. equal to $S^{-1}$. This
immigration prevents the collapse of the metacommunity to a single species,
since otherwise all species except one will face extinction by ecological drift
(speciation prevents this phenomenon to occur in \textcite{Hubbell2001}). The
local recruitment probability is consequently $Pr_{ix} = MS^{-1} + mP_{ix} +
(1-m-M)N_{ix}J_x^{-1}$.

The model is neutral because it assumes that the probabilities of local
recruitment, immigration and mortality events are all equal across species.
Demographic stochasticity is the source of variations in abundance, but larger
disturbances could be simulated as well, as long as they hit all species with
the same probability, independently of their abundance. The fundamental feature
of neutral dynamics is therefore the ecological drift, defined as population
changes emerging from neutraly stable population dynamics. It can be measured as
the variance between replicated time series of community dynamics
\parencite{Gravel2011a}. \textcite{Hubbell2001} provides a very comprehensive
analysis of the model, with specific attention to the effect of the different
parameters on drift (and consequently variance in abundance) and time to
extinction. Despite its simplicity, the neutral model is surprisingly rich in
the predictions it makes. \textcite{Bell2001} and \textcite{Hubbell2001}
analyzed the performance of neutral models to predict species abundance
distributions, the range-abundance relationship, spatial variation in abundance,
the species-area relationship, community turnover (beta-diversity) and
co-occurrence. Recent trophic neutral models were also found to predict
realistic ecological network structures \parencite{Canard2012}. Other than the
ecological equivalence assumption, one of the most criticized aspect of neutral
models is the realism of the speciation process and the required speciation
rates to sustain species richness \parencite{Ricklefs2003,Etienne2007}. Recent
neutral models with more credible speciation models
\parencite{Rosindell2010,Desjardins2012a} revealed the difficulty to maintain
diversity in neutral models over macro-evolutionary time scales. These models
nonetheless proposed interesting predictions on endemic species richness and
island biogeography \parencite{Rosindell2011,Desjardins2012b}.

@@

In this section we provide simple simulation results to illustrate the impact of
saptial contingencies on species distribution and coexistence. We consider four
different landscapes, illustrated at Fig. 1. with the outcome of simulations
using the neutral model. All of these networks have the same number of nodes,
but both different number of edges (e.g. dispersal routes between sampling
sites) and connectivity between nodes. We ask how these differences
in topological structures will shape the emerging properties of the community under the
scenarios represented by each metacommunity model. Our analysis is not
exhaustive, it is provided simply to illustrate the interaction between
metacommunity perspectives and landscape structures on $\alpha$, $\beta$ and
$\gamma$ diversity.

In Fig. 2, we present the species richness of each node of the network ($\alpha$
diversity), as a function of the centrality of the node, under different
assumptions of metacommunity dynamics and network structure. We scaled the
species richness by the maximal $\alpha$ diversity to facilitate comparison
between models. The model parameterization is responsible for differences in
both $\alpha$ and $\gamma$ diversity, meaning that only the shape of the
relationship between centrality and richness ought to be looked at. It appears
that both in the random geographical and tree networks, the path dynamics model has a
much more considerable variation in $\alpha$ diversity. However, in all
cases the $\alpha$ diversity increases with the node degree centrality, meaning
that nodes with more connections also host a more diverse community.
Eigen-centrality gave a ess clear-cut result, which can probably be
attributed to the fact that our networks are relatively small in size.
Eigen-centrality reports how well your neighbors are connected, and in graphs
with a short diameter (i.e. the two farthest points are not extremely far
apart), this measure might hold less information.
